---
title: "💻 🏦 <br>Code You Can Bank On"
subtitle: "Documentation"
author: "Amelia McNamara"
editor: source
format:
  revealjs: 
    theme: [dracula, extra.scss]
    slide-number: true
    height: 900
    width: 1600
execute:
  echo: true
  eval: false
---

# 🏷️ Document your package


## Levels of package documentation

- Metadata: The `DESCRIPTION` file -- "what's in this package?"

- Object documentation: for functions and data

- Package-level documentation

<!-- for each of the exported functions and datasets in the package, along with examples of usage -->

- Vignettes: Long form documentation

<!-- generally discussing how to use a number of functions from the - package together and/or how a package fits into a larger ecosystem of packages -->

# `DESCRIPTION`

## Metadata in `DESCRIPTION`

- Title: One line, title case, with no period. Fewer than 65 characters.

- Version

- for release: MAJOR.MINOR.PATCH version.

- for development version: MAJOR.MINOR.PATCH.9000

- Authors\@R:

- "aut" means author, "cre" means creator, "ctb" means contributor.


## Metadata in `DESCRIPTION`

- Description: One paragraph describing what the package does. Keep the width of the paragraph to 80 characters; indent subsequent lines with 4 spaces.

- License

- Encoding: How to encode text, use UTF-8 encoding.

- LazyData: Use true to lazy-load data sets in the package.

## Edit `DESCRIPTION` (your turn)

::: {.teach-left}

Open `DESCRIPTION` file to edit it:

- `Authors@R` field uses `options("usethis.description")` 

   - we set this in our `.Rprofile`, so go back to the intro slides if you missed this!
   - (Another possible issue is you didn't restart R!)

✏️ edit `Title` field. Should be one line, title case, with no period. Fewer than 65 characters.

💾 save `DESCRIPTION` file

:::

::: {.teach-right} 
```{r}
#| ref.label: noun-create-package
#| echo: false
```
:::

## Edit `DESCRIPTION`

::: small
``` default
Package: BLT
Title: Work with data on bacon, lettuce, and tomatoes
Version: 0.0.0.9000
Authors@R: 
    person("Amelia", "McNamara", , "amelia.mcnamara@stthomas.edu", role = c("aut", "cre"))
Description: What the package does (one paragraph).
License: `use_mit_license()`, `use_gpl3_license()` or friends to pick a
    license
Encoding: UTF-8
Roxygen: list(markdown = TRUE)
RoxygenNote: 7.3.3
```
:::


## Commit

Now would be a good time to commit your changes.

\



![Jason Long, CC BY 3.0 \<https://creativecommons.org/licenses/by/3.0\>, via Wikimedia Commons](https://upload.wikimedia.org/wikipedia/commons/3/3f/Git_icon.svg){fig-alt="Git icon" width="300"} 


## Create framework (your turn)

::: {.teach-left}

```r
use_roxygen_md()
```

- makes it **much** easier to create and maintain your function documentation

  - lets you use a markdown-like syntax


:::

::: {.teach-right} 
```{r}
#| ref.label: noun-framework
#| echo: false
```
:::




## Update `DESCRIPTION`

🎬 Edit the description

. . .

``` default
Description: This is a demo package for the "Code you can bank on" workshop at the Federal Reserve. It allows you to work with time series data, and example datasets on bacon, lettuce, and tomatoes. 
```

. . .

👀 The full stop matters!

## Add a license {.smaller}

::: {.teach-left}

```r
use_mit_license()
```

- lets the world know how it can use the stuff you create

  - CRAN requires you to specify a (valid) license
  - May be less relevant in your usecase!

- other choices include 
 
  - `use_apl2_license()`: Apache 2.0
  - `use_gpl3_license()`: GPL3

- uses `option("usethis.full_name")` or `option("devtools.name")` 
  - we set this in `.Rprofile`

:::


::: {.teach-right} 
```{r}
#| ref.label: noun-framework
#| echo: false
```
:::


## Commit and push

Now would be a good time to commit your changes.

\



![Jason Long, CC BY 3.0 \<https://creativecommons.org/licenses/by/3.0\>, via Wikimedia Commons](https://upload.wikimedia.org/wikipedia/commons/3/3f/Git_icon.svg){fig-alt="Git icon" width="300"} 

# Object documentation

## Develop (your turn)

::: {.teach-left}

Complete the documentation:

::: small
```r
#' Convenience function to make first-order autoregressive model
#'
#' Wrapper around forecast::Arima 
#
#' @param y a univariate time series of class `ts`
#'
#' @returns See the arima function in the stats package
#' @export
#'
#' @examples
#'  (AR1(LakeHuron))
#'
AR1 <- function(y) {

}
```
::: 

:::

::: {.teach-right} 
```{r}
#| ref.label: noun-develop
#| echo: false
```
:::

We're going to talk a lot more about documentation, but it's good to get in the habit early. 


## Object documentation

- Object documentation is what you see when you use `?` or `help()`

. . .

🎬 Try this now:

```{r}
?AR1
```

. . .

- Files are written in a special "R documentation" format: `.Rd`

- `.Rd` resembles LaTeX,

## `man/AR1.Rd`


\

::: extrasmall
``` default
% Generated by roxygen2: do not edit by hand
% Please edit documentation in R/AR1.R
\name{AR1}
\alias{AR1}
\title{Convenience function to make first-order autoregressive model}
\usage{
AR1(y)
}
\arguments{
\item{y}{a univariate time series of class \code{ts}}
}
\value{
See the arima function in the stats package
}
\description{
Wrapper around forecast::Arima
}
\examples{
 (AR1(LakeHuron))

}
```
:::

## Object documentation

🥳 We don't have to write it!!

. . .

- We write "roxygen comments" in the `.R` files

- These get turned into `.Rd` format

. . .

- Functions from **`roxygen2`** [@roxygen2] do the work but we can use **`devtools`** functions to call them.

## Roxygen comments

::: extrasmall
``` default
% Generated by roxygen2: do not edit by hand
% Please edit documentation in R/AR1.R
\name{AR1}
\alias{AR1}
\title{Convenience function to make first-order autoregressive model}
\usage{
AR1(y)
}
\arguments{
\item{y}{a univariate time series of class \code{ts}}
}
\value{
See the arima function in the stats package
}
\description{
Wrapper around forecast::Arima
}
\examples{
 (AR1(LakeHuron))

}
```
:::

- the `#'` indicates it is a roxygen comment

## Object documentation workflow

- Add roxygen comments to your `.R` files.

. . .

- Run `devtools::document()` to convert roxygen comments to `.Rd` files.

. . .

- Load the package with `devtools::load_all()`

. . .

- Preview documentation with `?`

. . .

- Repeat until the documentation looks the way you want.

## Edit function documentation

🎬 Open `AR1.R`

. . .

```{r}
usethis::use_r("AR1")
```


## Roxygen comments and tags

- `#'` indicates it is a roxygen comment

- `@` indicates a roxygen tag

  - `@param arg` --- describe the inputs
  - `@examples`--- show how the function works
  - `@seealso` --- point out related functions
  - `@return` --- describe the outputs
  - `@export` --- is this a user visible function

## Document your function

🎬 Give your function a title and a brief description

\

🎬 Define the parameter(s)

\

🎬 Describe what the function returns

## Our answer

::: small
``` default
#' Convenience function to make first-order autoregressive model
#'
#' Wrapper around forecast::Arima
#
#' @param y a univariate time series of class `ts`
#'
#' @returns See the arima function in the stats package
#' @export
#'
#' @examples
#'  (AR1(LakeHuron))
#'

```
:::

. . .

\

🎬 Save `AR1.R`

## Build documentation

🎬 Run `devtools::document()` to turn roxygen comments to `.Rd`

```{r}
devtools::document()
```

. . .

``` default
ℹ Updating BLT documentation
ℹ Loading BLT
Writing NAMESPACE
Writing AR1.Rd
Warning message:
[matches.R:13] @examples requires a value 
```

. . .

We will that fix later!

## Take a look 1

``` default
├── DESCRIPTION
├── LICENSE
├── LICENSE.md
├── man
│   └── uss_make_matches.Rd 
├── NAMESPACE
├── R
│   └── matches.R
└── BLT.Rproj
```

- `NAMESPACE` now lists the exported function

- `man/uss_make_matches.Rd` has been written

## Take a look 2

🎬 Load the package : `devtools::load_all()`

\

🎬 Preview the documentation with:

```{r}
?AR1
```

. . .

\

😮 🥳

## Fix warning

``` default
Warning message:
[matches.R:13] @examples requires a value 
```

. . .

We need to add a usage example!

🎬 Under `@examples`, add one example for using your function

. . .

``` default
#' @examples
#' uss_make_matches(engsoccerdata::spain, "Spain")
```

## Look again

🎬 Save `matches.R`, run `devtools::document()` followed by `devtools::load_all()`

\

. . .

🎬 Preview the documentation with `?uss_make_matches` and edit if needed

## Package-level help page

🎬 What happens if you do:

```{r}
?BLT
```

. . .

``` default
No documentation for 'BLT' in specified packages and libraries:
you could try '??BLT'
```

## Package-level help page

We can fix this with `use_package_doc()`\

. . .

🎬 Add a package-level help page:

```{r}
usethis::use_package_doc()
```

. . .

\

``` default
✔ Writing 'R/BLT-package.R'
• Modify 'R/BLT-package.R'
```

. . .

\

🎬 Run `devtools::document()` then `devtools::load_all()` followed by `?BLT`

## Commit and push

Now would be a good time to commit your changes 

![Jason Long, CC BY 3.0 \<https://creativecommons.org/licenses/by/3.0\>, via Wikimedia Commons](https://upload.wikimedia.org/wikipedia/commons/3/3f/Git_icon.svg){fig-alt="Git icon" width="300"} 

## `devtools::check()`

This is a good time to run R CMD check to ensure our package is in full working order.

. . .

🎬 Use **`devtools`** to run R CMD check

. . .

```{r}
devtools::check()
```

## Documenting data

Data gets documented in the data.R file. How can we create and open this file for editing?

. . .

```r
use_r("data")
```

Unfortunately, the trick to insert Roxygen skeleton doesn't work with data, so we have to do the documentation more manually. 

## Documenting data

Here's what the `who` documentation in `tidyr` looks like

::: small
```
#' World Health Organization TB data
#'
#' A subset of data from the World Health Organization Global Tuberculosis
#' Report ...
#'
#' @format ## `who`
#' A data frame with 7,240 rows and 60 columns:
#' \describe{
#'   \item{country}{Country name}
#'   \item{iso2, iso3}{2 & 3 letter ISO country codes}
#'   \item{year}{Year}
#'   ...
#' }
#' @source <https://www.who.int/teams/global-tuberculosis-programme/data>
"who"
```

:::

Can you write some documentation for the `bacon` data?

## Documenting data (your turn)


::: small
```
#' FRED bacon price data
#'
#' A subset of data on the average price of sliced bacon in the US.
#' Data is from 2016 to 2026
#'
#' @format ## `bacon`
#' A data frame with 119 rows and 5 columns:
#' \describe{
#'   \item{date}{Date of observation}
#'   \item{series_id}{APU0000704111, the series ID for the series}
#'   \item{value}{the average price of bacon at that date}
#'   \item{real_timestart, realtime_end}{the date the data was downloaded}
#' }
#' @source <https://fred.stlouisfed.org/series/APU0000704111>
"bacon"
```

:::

## 

::: extrasmall
``` default
── R CMD check results ─────────────────────────────────── BLT 0.0.0.9000 ────
Duration: 40.4s

❯ checking dependencies in R code ... WARNING
  '::' or ':::' imports not declared from:
    'dplyr' 'tibble'

❯ checking for unstated dependencies in examples ... WARNING
  '::' or ':::' import not declared from: 'engsoccerdata'

❯ checking R code for possible problems ... NOTE
  uss_make_matches: no visible binding for global variable 'tier'
  uss_make_matches: no visible binding for global variable 'Season'
  uss_make_matches: no visible binding for global variable 'Date'
  uss_make_matches: no visible binding for global variable 'home'
  uss_make_matches: no visible binding for global variable 'visitor'
  uss_make_matches: no visible binding for global variable 'hgoal'
  uss_make_matches: no visible binding for global variable 'vgoal'
  Undefined global functions or variables:
    Date Season hgoal home tier vgoal visitor

0 errors ✔ | 2 warnings ✖ | 1 note ✖
```
:::

\

We've gained a warning!

\

Our warnings are about Package dependencies



# Package dependencies

## Package dependencies

Our warnings are because we have used packages that we have not declared officially.

\

We need to document our package dependencies

\

. . .

Our users, and the package installation machinery, need to know what our package depends on before installing

## Package dependencies

Levels of dependency

- **Imports:** must be installed for your package to work. If they're not, they will get installed.

. . .

- **Suggests:** used by your package, but not required. Might provide data for examples, to run tests, build vignettes.

. . .

- **Depends:** Avoid where possible. When your package requires a specific version of R, e.g. `Depends: R (>= 3.4.0)`. Think critically: downstream effects on packages that depend on your package.

## 

``` default
❯ checking dependencies in R code ... WARNING
  '::' or ':::' imports not declared from:
    'dplyr' 'tibble'

❯ checking for unstated dependencies in examples ... WARNING
  '::' or ':::' import not declared from: 'engsoccerdata'
```

## Package dependencies

**`usethis`** again!

. . .

`use_package(package, type = "Imports")`

- Type -- one of "Imports", "Depends", "Suggests", "Enhances", or "LinkingTo"

. . .

- The default is "Imports"

## `usethis::use_package()`

🎬 Use `usethis::use_package()` to add the **`dplyr`** package to `Imports`

. . .

```{r}
usethis::use_package("dplyr")
```

\

. . .

``` default
✔ Adding 'dplyr' to Imports field in DESCRIPTION
• Refer to functions with `dplyr::fun()`
```

\

Look how your `DESCRIPTION` file changed!

## `usethis::use_package()`

Note: we get reminded to "Refer to functions with `dplyr::fun()`"

 

. . .

That is, we do NOT use `library(dplyr)` to make functions available to our package.

## `devtools::check()`

🎬 Run `devtools::check()` on your package again

## 

``` default
── R CMD check results ─────────── BLT 0.0.0.9000 ────
Duration: 24s

❯ checking dependencies in R code ... WARNING
  '::' or ':::' import not declared from: 'tibble'

❯ checking for unstated dependencies in examples ... WARNING
  '::' or ':::' import not declared from: 'engsoccerdata'

❯ checking R code for possible problems ... NOTE
  uss_make_matches: no visible binding for global variable 'tier'
  uss_make_matches: no visible binding for global variable 'Season'
  uss_make_matches: no visible binding for global variable 'Date'
  uss_make_matches: no visible binding for global variable 'home'
  uss_make_matches: no visible binding for global variable 'visitor'
  uss_make_matches: no visible binding for global variable 'hgoal'
  uss_make_matches: no visible binding for global variable 'vgoal'
  Undefined global functions or variables:
    Date Season hgoal home tier vgoal visitor

0 errors ✔ | 2 warnings ✖ | 1 note ✖
```

## Imports

Three types, according to how much your are using the imported package.

. . .

If you are using....

- a few functions from another package: put package name in the `Imports:` field of the `DESCRIPTION` file and call the function(s) explicitly using `::`, e.g., `pkg::fun()`.

. . .

- a function repeatedly, uses you can avoid `::` by importing the function with `@importFrom pkg fun` in `BLT-package.R`.

. . .

- many functions from another package repeatedly: `@import package`. But beware, less transparent, higher likelihood of conflicting function names.

## Imports

We will use `tibble::tible()` multiple times.

. . .

Instead of using:

```{r}
usethis::use_package("tibble")
```

. . .

we can use

```{r}
usethis::use_import_from("tibble", "tibble")
```

. . .

which will:

- to add the package to the `DESCRIPTION` imports
- add `@importFrom tibble tibble` to `BLT-package.R` and `NAMESPACE`
- mean we can avoid `::`

## `@importFrom pkg fun`

🎬 Import `tibble()` from **`tibble`**:

```{r}
usethis::use_import_from("tibble", "tibble")
```

. . .

``` default
✔ Adding 'tibble' to Imports field in DESCRIPTION
✔ Adding '@importFrom tibble tibble' to 'R/BLT-package.R'
✔ Writing 'NAMESPACE'
✔ Loading BLT
```

## Suggests

- Are used by your package, but not required for it to work

- Might provide data for examples

. . .

In our case:

`matches.R`

```         
#' @examples
#' uss_make_matches(engsoccerdata::spain, "Spain")
```

## Suggests

🎬 Add **`engsoccerdata`**

. . .

```{r}
usethis::use_package("engsoccerdata",
                     type = "Suggests")
```

 

``` default
✔ Adding 'engsoccerdata' to Suggests field in DESCRIPTION
• Use `requireNamespace("engsoccerdata", quietly = TRUE)` to test if package is installed
• Then directly refer to functions with `engsoccerdata::fun()`
```

## Suggests

- Packages listed in Suggests are not automatically installed along with your package.

- This means that you can't assume the package is available


# 🏷️Yee haw! 🏷️ <BR> Your package has documentation!

## Commit and push

Now would be a good time to commit your changes 

![Jason Long, CC BY 3.0 \<https://creativecommons.org/licenses/by/3.0\>, via Wikimedia Commons](https://upload.wikimedia.org/wikipedia/commons/3/3f/Git_icon.svg){fig-alt="Git icon" width="300"} 

## Summary

::: small
- A package has metadata and object documentation (essential), vignettes and **`pkgdown`** sites (optional)
- Metadata is in the `DESCRIPTION`
  - some fields need manual editing
  - some fields are appropriately edited by **`devtools`** workflow functions
- Objects like functions and data are documented with roxygen comments then turned into documentation with `devtools::document()`
  - roxygen comments are indicated with `#'`
  - roxygen tags start with `@`
:::

## Summary continued

::: small
- Package dependencies need to be documented
  - package dependencies are added with `use_package()`
  - functions from dependencies are called with `pkg::function()`
  - we don't use `library()`
- `cffr::cff_write()` will write a citation file from the description file.
:::

