---
title: "Documentation"
editor: source
format:
  revealjs: 
    theme: [dracula, extra.scss]
    slide-number: true
    height: 900
    width: 1600
execute:
  echo: true
  eval: false
---

# 🏷️ Document your package


## Levels of package documentation

- Metadata: The `DESCRIPTION` file -- "what's in this package?"

- Object documentation: for functions and data

- Package-level documentation

<!-- for each of the exported functions and datasets in the package, along with examples of usage -->

- Vignettes: Long form documentation

<!-- generally discussing how to use a number of functions from the - package together and/or how a package fits into a larger ecosystem of packages -->

# `DESCRIPTION`

## Metadata in `DESCRIPTION`

- Title: One line, title case, with no period. Fewer than 65 characters.

- Version

- for release: MAJOR.MINOR.PATCH version.

- for development version: MAJOR.MINOR.PATCH.9000

- Authors\@R:

- "aut" means author, "cre" means creator, "ctb" means contributor.

## Update `DESCRIPTION`

🎬 Edit the title

\

🎬 You should find that your information as an author is already included because you edited your `.RProfile` in intro

## Update `DESCRIPTION`

::: small
``` default
Package: BLT
Title: Work with data on bacon, lettuce, and tomatoes
Version: 0.0.0.9000
Authors@R: 
    person("Amelia", "McNamara",
          email = "amelia.mcnamara@stthomas.edu",
          role = c("aut", "cre"))
Description: What the package does (one paragraph).
License: MIT + file LICENSE
Encoding: UTF-8
Roxygen: list(markdown = TRUE)
RoxygenNote: 7.2.0
```
:::

## Metadata in `DESCRIPTION`

- Description: One paragraph describing what the package does. Keep the width of the paragraph to 80 characters; indent subsequent lines with 4 spaces.

- License

- Encoding: How to encode text, use UTF-8 encoding.

- LazyData: Use true to lazy-load data sets in the package.

## Update `DESCRIPTION`

🎬 Edit the description

. . .

``` default
Description: This is a demo package for the "Code you can bank on" workshop at the Federal Reserve. It allows you to work with time series data.
```

. . .

👀 The full stop matters!

## Commit and push

Now would be a good time to commit your changes.

\



![Jason Long, CC BY 3.0 \<https://creativecommons.org/licenses/by/3.0\>, via Wikimedia Commons](https://upload.wikimedia.org/wikipedia/commons/3/3f/Git_icon.svg){fig-alt="Git icon" width="300"} 

# Object documentation

## Object documentation

- Object documentation is what you see when you use `?` or `help()`

. . .

🎬 Try this now:

```{r}
?AR1
```

. . .

- Files are written in a special "R documentation" format: `.Rd`

- `.Rd` resembles LaTeX,

## `man/AR1.Rd`


\

::: extrasmall
``` default
% Generated by roxygen2: do not edit by hand
% Please edit documentation in R/fbind.R
\name{AR1}
\alias{AR1}
\title{Convenience function to make first-order autoregressive model}
\usage{
AR1(y)
}
\arguments{
\item{y}{a univariate time series of class \code{ts}}

\item{...}{Additional arguments to be passed to \href{arima}{stats::arima()}.}
}
\value{
See the \code{\link[stats:arima]{arima}} function in the stats package
}
\description{
Convenience function to make first-order autoregressive model
}
\examples{
(AR1(LakeHuron))
```
:::

## Object documentation

🥳 We don't have to write it!!

. . .

- We write "roxygen comments" in the `.R` files

- These get turned into `.Rd` format

. . .

- Functions from **`roxygen2`** [@roxygen2] do the work but we can use **`devtools`** functions to call them

## Roxygen comments

::: extrasmall
``` default
#' Convenience function to make first-order autoregressive model
#'
#' @param y a univariate time series of class `ts`
#' @param ... Additional arguments to be passed to [`arima`][stats::arima()]
#'
#' @returns See the [`arima`][stats::arima()] function in the stats package
#' @export
#'
#' @examples
#' (AR1(LakeHuron))
AR1 <- function(y) {
  forecast::Arima(y, order = c(1, 0, 0), include.constant = TRUE)

}
```
:::

- the `#'` indicates it is a roxygen comment

## Object documentation workflow

- Add roxygen comments to your `.R` files.

. . .

- Run `devtools::document()` to convert roxygen comments to `.Rd` files.

. . .

- Load the package with `devtools::load_all()`

. . .

- Preview documentation with `?`

. . .

- Repeat until the documentation looks the way you want.

## Document your function

🎬 Open `BLT.R`

. . .

```{r}
usethis::use_r("BLT")
```

\
. . .

🎬 Put your cursor anywhere in the function and do Code \| Insert Roxygen Documentation (Ctrl-Alt-Shift-R)

## 

::: small
``` default
#' Title
#'
#' @param data_engsoc 
#' @param country 
#'
#' @return
#' @export
#'
#' @examples
uss_make_matches <- function(data_engsoc, country) {
  result <-
    data_engsoc |>
    tibble::as_tibble() |>
    dplyr::transmute(
      country = as.character(country),
      tier = as.integer(tier),
      season = as.integer(Season),
      date = as.Date(Date),
      home = as.character(home),
      visitor = as.character(visitor),
      goals_home = as.integer(hgoal),
      goals_visitor = as.integer(vgoal)
    )
  result
}
```
:::

## Roxygen comments and tags

- `#'` indicates it is a roxygen comment

- `@` indicates a roxygen tag

  - `@param arg` --- describe the inputs
  - `@examples`--- show how the function works
  - `@seealso` --- point out related functions
  - `@return` --- describe the outputs
  - `@export` --- is this a user visible function

## Document your function

🎬 Give your function a title and a brief description

\

🎬 Define the parameter

\

🎬 Describe what the function returns

## Our answer

::: small
``` default
#' Make a standard league-play tibble
#'
#' Given a league-play data frame from {engsoccer}, returns a tibble with
#' standardised column-names and types, e.g. `date` is a `Date`.
#'
#' @param data_engsoc  obtained from {engsoccerdata}.
#' @param country `character` scalar, specifies the league.
#'
#' @return a tibble with columns `country`, `date`, `season`, `tier`, `home`,
#'    `visitor`, `goals_home`, `goals_visitor`.
#' @export
#'
#' @examples
```
:::

. . .

\

🎬 Save `matches.R`

## Build documentation

🎬 Run `devtools::document()` to turn roxygen comments to `.Rd`

```{r}
devtools::document()
```

. . .

``` default
ℹ Updating ussie documentation
ℹ Loading ussie
Writing NAMESPACE
Writing uss_make_matches.Rd
Warning message:
[matches.R:13] @examples requires a value 
```

. . .

We will that fix later!

## Take a look 1

``` default
├── DESCRIPTION
├── LICENSE
├── LICENSE.md
├── man
│   └── uss_make_matches.Rd 
├── NAMESPACE
├── R
│   └── matches.R
└── ussie.Rproj
```

- `NAMESPACE` now lists the exported function

- `man/uss_make_matches.Rd` has been written

## Take a look 2

🎬 Load the package : `devtools::load_all()`

\

🎬 Preview the documentation with:

```{r}
?uss_make_matches
```

. . .

\

😮 🥳

## Fix warning

``` default
Warning message:
[matches.R:13] @examples requires a value 
```

. . .

We need to add a usage example!

🎬 Under `@examples`, add one example for using your function

. . .

``` default
#' @examples
#' uss_make_matches(engsoccerdata::spain, "Spain")
```

## Look again

🎬 Save `matches.R`, run `devtools::document()` followed by `devtools::load_all()`

\

. . .

🎬 Preview the documentation with `?uss_make_matches` and edit if needed

## Package-level help page

🎬 What happens if you do:

```{r}
?ussie
```

. . .

``` default
No documentation for 'ussie' in specified packages and libraries:
you could try '??ussie'
```

## Package-level help page

We can fix this with `use_package_doc()`\

. . .

🎬 Add a package-level help page:

```{r}
usethis::use_package_doc()
```

. . .

\

``` default
✔ Writing 'R/ussie-package.R'
• Modify 'R/ussie-package.R'
```

. . .

\

🎬 Run `devtools::document()` then `devtools::load_all()` followed by `?ussie`

## Commit and push

Now would be a good time to commit your changes 

![Jason Long, CC BY 3.0 \<https://creativecommons.org/licenses/by/3.0\>, via Wikimedia Commons](https://upload.wikimedia.org/wikipedia/commons/3/3f/Git_icon.svg){fig-alt="Git icon" width="300"} 

## `devtools::check()`

This is a good time to run R CMD check to ensure our package is in full working order.

. . .

🎬 Use **`devtools`** to run R CMD check

. . .

```{r}
devtools::check()
```

## 

::: extrasmall
``` default
── R CMD check results ─────────────────────────────────── ussie 0.0.0.9000 ────
Duration: 40.4s

❯ checking dependencies in R code ... WARNING
  '::' or ':::' imports not declared from:
    'dplyr' 'tibble'

❯ checking for unstated dependencies in examples ... WARNING
  '::' or ':::' import not declared from: 'engsoccerdata'

❯ checking R code for possible problems ... NOTE
  uss_make_matches: no visible binding for global variable 'tier'
  uss_make_matches: no visible binding for global variable 'Season'
  uss_make_matches: no visible binding for global variable 'Date'
  uss_make_matches: no visible binding for global variable 'home'
  uss_make_matches: no visible binding for global variable 'visitor'
  uss_make_matches: no visible binding for global variable 'hgoal'
  uss_make_matches: no visible binding for global variable 'vgoal'
  Undefined global functions or variables:
    Date Season hgoal home tier vgoal visitor

0 errors ✔ | 2 warnings ✖ | 1 note ✖
```
:::

\

We've gained a warning!

\

Our warnings are about Package dependencies

# Package dependencies

## Package dependencies

Our warnings are because we have used packages that we have not declared officially.

\

We need to document our package dependencies

\

. . .

Our users, and the package installation machinery, need to know what our package depends on before installing

## Package dependencies

Levels of dependency

- **Imports:** must be installed for your package to work. If they're not, they will get installed.

. . .

- **Suggests:** used by your package, but not required. Might provide data for examples, to run tests, build vignettes.

. . .

- **Depends:** Avoid where possible. When your package requires a specific version of R, e.g. `Depends: R (>= 3.4.0)`. Think critically: downstream effects on packages that depend on your package.

## 

``` default
❯ checking dependencies in R code ... WARNING
  '::' or ':::' imports not declared from:
    'dplyr' 'tibble'

❯ checking for unstated dependencies in examples ... WARNING
  '::' or ':::' import not declared from: 'engsoccerdata'
```

## Package dependencies

**`usethis`** again!

. . .

`use_package(package, type = "Imports")`

- Type -- one of "Imports", "Depends", "Suggests", "Enhances", or "LinkingTo"

. . .

- The default is "Imports"

## `usethis::use_package()`

🎬 Use `usethis::use_package()` to add the **`dplyr`** package to `Imports`

. . .

```{r}
usethis::use_package("dplyr")
```

\

. . .

``` default
✔ Adding 'dplyr' to Imports field in DESCRIPTION
• Refer to functions with `dplyr::fun()`
```

\

Look how your `DESCRIPTION` file changed!

## `usethis::use_package()`

Note: we get reminded to "Refer to functions with `dplyr::fun()`"

 

. . .

That is, we do NOT use `library(dplyr)` to make functions available to our package.

## `devtools::check()`

🎬 Run `devtools::check()` on your package again

## 

``` default
── R CMD check results ─────────── ussie 0.0.0.9000 ────
Duration: 24s

❯ checking dependencies in R code ... WARNING
  '::' or ':::' import not declared from: 'tibble'

❯ checking for unstated dependencies in examples ... WARNING
  '::' or ':::' import not declared from: 'engsoccerdata'

❯ checking R code for possible problems ... NOTE
  uss_make_matches: no visible binding for global variable 'tier'
  uss_make_matches: no visible binding for global variable 'Season'
  uss_make_matches: no visible binding for global variable 'Date'
  uss_make_matches: no visible binding for global variable 'home'
  uss_make_matches: no visible binding for global variable 'visitor'
  uss_make_matches: no visible binding for global variable 'hgoal'
  uss_make_matches: no visible binding for global variable 'vgoal'
  Undefined global functions or variables:
    Date Season hgoal home tier vgoal visitor

0 errors ✔ | 2 warnings ✖ | 1 note ✖
```

## Imports

Three types, according to how much your are using the imported package.

. . .

If you are using....

- a few functions from another package: put package name in the `Imports:` field of the `DESCRIPTION` file and call the function(s) explicitly using `::`, e.g., `pkg::fun()`.

. . .

- a function repeatedly, uses you can avoid `::` by importing the function with `@importFrom pkg fun` in `ussie-package.R`.

. . .

- many functions from another package repeatedly: `@import package`. But beware, less transparent, higher likelihood of conflicting function names.

## Imports

We will use `tibble::tible()` multiple times.

. . .

Instead of using:

```{r}
usethis::use_package("tibble")
```

. . .

we can use

```{r}
usethis::use_import_from("tibble", "tibble")
```

. . .

which will:

- to add the package to the `DESCRIPTION` imports
- add `@importFrom tibble tibble` to `ussie-package.R` and `NAMESPACE`
- mean we can avoid `::`

## `@importFrom pkg fun`

🎬 Import `tibble()` from **`tibble`**:

```{r}
usethis::use_import_from("tibble", "tibble")
```

. . .

``` default
✔ Adding 'tibble' to Imports field in DESCRIPTION
✔ Adding '@importFrom tibble tibble' to 'R/ussie-package.R'
✔ Writing 'NAMESPACE'
✔ Loading ussie
```

## Suggests

- Are used by your package, but not required for it to work

- Might provide data for examples

. . .

In our case:

`matches.R`

```         
#' @examples
#' uss_make_matches(engsoccerdata::spain, "Spain")
```

## Suggests

🎬 Add **`engsoccerdata`**

. . .

```{r}
usethis::use_package("engsoccerdata",
                     type = "Suggests")
```

 

``` default
✔ Adding 'engsoccerdata' to Suggests field in DESCRIPTION
• Use `requireNamespace("engsoccerdata", quietly = TRUE)` to test if package is installed
• Then directly refer to functions with `engsoccerdata::fun()`
```

## Suggests

- Packages listed in Suggests are not automatically installed along with your package.

- This means that you can't assume the package is available


# 🏷️Yee haw! 🏷️ <BR> Your package has documentation!

## Commit and push

Now would be a good time to commit your changes 

![Jason Long, CC BY 3.0 \<https://creativecommons.org/licenses/by/3.0\>, via Wikimedia Commons](https://upload.wikimedia.org/wikipedia/commons/3/3f/Git_icon.svg){fig-alt="Git icon" width="300"} 

## Summary

::: small
- A package has metadata and object documentation (essential), vignettes and **`pkgdown`** sites (optional)
- Metadata is in the `DESCRIPTION`
  - some fields need manual editing
  - some fields are appropriately edited by **`devtools`** workflow functions
- Objects like functions and data are documented with roxygen comments then turned into documentation with `devtools::document()`
  - roxygen comments are indicated with `#'`
  - roxygen tags start with `@`
:::

## Summary continued

::: small
- Package dependencies need to be documented
  - package dependencies are added with `use_package()`
  - functions from dependencies are called with `pkg::function()`
  - we don't use `library()`
- `cffr::cff_write()` will write a citation file from the description file.
:::

