---
title: "💻 🏦 <br>Code You Can Bank On"
subtitle: "Side effects"
author: "Amelia McNamara"
editor: source
format:
  revealjs: 
    theme: [dracula, extra.scss]
    slide-number: true
    footer: <https://bit.ly/CYCBO>   
    height: 900
    width: 1600
execute:
  echo: true
  eval: false
---
## Pure function

A function where:

 - the return value depends *only* on argument values

 - only change is the return value

. . .

Examples:

. . .
 
```r
function(x, y) {
  x + y
}
```

. . .
 
```r
cos(pi)
```

## Side effects

A function where:

 - the return value can depend on "the outside universe"

 - there is a change in the "the outside universe"
 
. . .

Examples:

. . .

```r
readr::read_csv("myfile.csv")
```

. . .

```r
runif(1)
```

## Why is this important?

. . . 

- Pure functions easier to test than functions with side effects.

. . . 

- Side effects (interactions with universe) take time (Shiny).

. . . 

- Functions with side effects should document the effects.

. . . 

- Side effects are not inherently bad (we *do* need to write to the file system),
  but they need extra care.
  
  

## Errors

- The most common type of side-effect is the **error condition**. 

. . .

- Sometimes, error messages can be cryptic: 

  ```r
  seq[10]
  ```

  ```
  Error in seq[10] : object of type 'closure' is not subsettable
  ```

. . . 


- You can write error messages that make things clearer for: 
  - developers who call your functions
  - end users
  
  
## Making errors

There are many ways to create error messages for your R functions. 

In base R, you might use `if()`/`else()` statements with functions like `message()`, `warning()` and `stop()` inside, or a `stopifnot()`. 

Of course, the folks who make the tidyverse have also been thinking about error messages. 

In 2013, Hadley wickham created `{assertthat}` as a drop-in replacement for `stopifnot()`. People still use `assertthat`! But, the tidyverse always moves on.

The new coolness is `{cli}`, which was released in 2017. 

## {cli}

The `{cli}` package is technically for command line interfaces, so it is more general than just error messages. But, it gets used in many situations where `assertthat` might have been previously. 

I went looking for a repo that had a mixture of `assertthat` and `cli`, and found the [{fishmecher}](https://github.com/tytell/fishmechr/tree/main) package. Check out [center_of_mass.R](https://github.com/tytell/fishmechr/blob/main/R/center_of_mass.R). 

(Actually, I don't think they're using `cli` quite right.)


## Creating an error condition 

An effective error condition has:

1. **predicate** (logical expression used to identify condition)
2. clear **message** for end user
3. **class name** for developer
4. **more information** for developer

Previous error-handling code had 1, and sometimes 2. But `cli` is really adding in the 3 and 4. 

## Using `cli::cli_abort()`

{cli} package

```r
# predicate
if (y > 3) {
  cli::cli_abort(
    # message for end user
    c(
      "{.var y} cannot be greater than 3.", 
      x = "{.var y} is {.val {y}}."
    ),
    # class name
    class = "BLT_error_threshold",  
    # more information
    y = y     
  )
}
```

## Predicate

Prefer:

 - simpler predicates and more error-conditions
 
Over:

 - complex predicates and fewer error-conditions
 
. . .

Finding simplest set of predicates is just as challenging as finding the "right" names for functions and arguments.

## Message

Content:

- how did we violate the predicate?

. . . 

Formatting:

- {cli} provides powerful formatting tools:

  - use curly-braces and a tag, e.g. `{.var y}` 
  - use more curly-braces to interpolate, e.g. `{.val {y}}`
  
- see [cli inline-markup](https://cli.r-lib.org/reference/inline-markup.html) for more details.

## Class name

- A developer, calling your function, can use the `class` name to handle the error, if they want.

- Convention: 
  
  - `"{package}_error_{description}"`

## Additional information

This "stuff" is also available to an error handler.

- Provide the data that went into the predicate.

- Provide name of variable, e.g. `y = y`.

- Avoid reserved names: `message`, `class`, `call`, `body`, `footer`, `trace`, `parent`, `use_cli_format`.

## Validation

If there will be an error, surface it quickly.

Validating the arguments to a function is one way to do this.

. . .

For example:

 - is *this* a data frame?
 - does this data frame have *these* columns?

. . . 
 
Questions like these can be generalized into functions:

 - throw an error if you need to.
 - otherwise, return **data** argument invisibly.
 
## Validate string-values 

You may be familiar with `match.arg()`:

 - uses "magic" to compare value (if any) to default
 - argument default is vector of strings
   - if value among defaults, return it - otherwise **error**
   - if no value, return *first* value in default

. . .

`rlang::arg_match()` does the same thing, but:

  - partial match triggers error
  - error messages conform to tidyverse standards


   
## Validator functions

::: extrasmall
```r
validate_data_frame <- function(.data, call = rlang::caller_env()) {
  # predicate
  if (!is.data.frame(.data)) {
    cli::cli_abort(
      # information for user
      c(
        "Must supply a data frame",
        x = "You have supplied a {.cls {class(.data)}}."
      ),
      # class information for developers
      class = "BLT_error_data",
      # other information for developers
      class_data = class(.data),
      # tell user where we are calling from
      call = call
    )
  }
  
  invisible(.data)
}
```
:::

## Validator functions

::: extrasmall
```r
validate_cols <- function(.data, cols_req, call = rlang::caller_env()) {

  cols_data <- names(.data)
  cols_missing <- cols_req[!(cols_req %in% cols_data)]

  # predicate
  if (length(cols_missing) > 0) {

    # helper function to format quantities
    # - see https://cli.r-lib.org/articles/pluralization.html
    qlen <- function(x) cli::qty(length(x))

    cli::cli_abort(
      # information for user
      c(
        "Data frame needs {qlen(cols_req)} column{?s}: {.var {cols_req}}",
        i = "Has {qlen(cols_data)} column{?s}: {.var {cols_data}}",
        x = "Missing {qlen(cols_missing)} column{?s}: {.var {cols_missing}}"
      ),
      # class information for developers
      class = "ussie_error_cols",
      # other information for developers
      cols_req = cols_req,
      cols_data = cols_data,
      # tell user where we are calling from
      call = call
    )
  }

  invisible(.data)
}
```
:::


## Build error constructor {.smaller}

- `usethis::use_package("cli")`

- review `validate_data_frame()` (`call` argument)

- build constructor for `validate_cols()`:
  - something like `{.field {cols}}` may be useful

- add validator-functions to `perc_missing_tidy()`




## Make a function with side effects

So far, our functions have been pure. Let's make one that is not. For this function, I want to write a wrapper around `imputeTS::ggplot_na_distribution()`. This function has the name `ggplot` in the title, but it doesn't work like a normal `ggplot()` function (taking the data as the first argument, then `aes`thetics). 

. . .

Do you remember the `usethis` function that will open up a .R file for us to write `BLT_ggplot_na_distribution.R`?

. . . 

```{r}
usethis::use_r()
```


Let's start with adding a function skeleton,

. . .

```r
BLT_ggplot_na_distribution <- function(data, mapping, ...){

}
```

. . . 

And then write some documentation **first**, as is better practice. 

## Make a function with side effects

::: extrasmall
```r
#' ggplot to visualization the distribution of missing values
#'
#' Wrapper around [imputeTS::ggplot_na_distribution]
#'
#' @param data Default dataset to use for a plot
#' @param variable The variable to plot
#' @param ... Other arguments passed on to methods
#'
#' @returns a ggplot object
#' @export
#'
#' @examples
#' tomatoes |>
#' BLT_ggplot_na_distribution(value)
BLT_ggplot_na_distribution <- function(data, variable, ...){

}
```
:::


## Make a function with side effects

::: extrasmall
```r
#' ggplot to visualization the distribution of missing values
#'
#' Wrapper around [imputeTS::ggplot_na_distribution]
#'
#' @param data Default dataset to use for a plot
#' @param variable The variable to plot
#' @param ... Other arguments passed on to methods
#'
#' @returns a ggplot object
#' @export
#'
#' @examples
#' tomatoes |>
#' BLT_ggplot_na_distribution(value)
BLT_ggplot_na_distribution <- function(data, variable, ...){
  var <- rlang::enquo(variable)
  var_eval <- rlang::eval_tidy(var, data = data)
  imputeTS::ggplot_na_distribution(var_eval)
}
```
:::


## Practical advice

Try to separate tasks into pure functions and side effects:

- easier to test the pure functions and side effects separately

- use *these* functions in higher-level functions



## Snapshot tests

Designed for capturing side-effects:

 - error messages
 - `dplyr::glimpse()` is a useful shortcut
 
. . .  
 
Be careful about accepting changes (don't just accept).

Can be temperamental - not run on CRAN.

We will go through, using examples.


## Managing global state

Leave no footprints.

Leave the global state how you found it, avoid surprises later:

 - packages loaded
 - also: options, environment variables 

. . .

Loading {devtools} in `.Rprofile` changes the global state.

. . .

When we hit the "Knit" button, or the Quarto "Render" button:

  - it runs in a *new* R session
  - does not execute user's `.Rprofile`
  
## Using "self-removing footprints"

The {withr} package [@withr] gives us tools to:

 - modify global state
 - specify when to reverse the modification

. . . 
 
Useful family of functions:

 - `withr::local_*()`
 - changes the state back when a scope is exited
 - if called within a function, *normally* when the function exits

## When could this be useful?

CRAN is (rightly) particular about "leave no footprints".

You may need:

 - `withr::local_options()`: change an option
 - `withr::local_dir()`: change the working directory
 - `withr::local_tempfile()`: path to a temporary file
 
. . . 

Useful in `testthat` code and in `R` code.

## Interactive example

```{r}
write_read_vanish <- function(x) {
  
  tempfile <- withr::local_tempfile(fileext = ".rds")
  
  saveRDS(x, file = tempfile)
  xnew <- readRDS(tempfile)
  
  print(
    glue::glue("'{tempfile}' contained {xnew}.")
  )
  
  invisible(xnew)
}
```

## Another interactive example

```r
not_runif <- function(n, min = 0, max = 1) {

  withr::local_seed(314)
  
  runif(n, min = min, max = max)
}
```

## Example

- `usethis::use_test("validate")`
- change: 

  ```r
  expect_identical(error_condition$cols_req, "foo")
  ```
  
- to: 

  ```r
  expect_identical(error_condition$cols_re, "foo")
  ```

- rerun tests 🤔

. . . 

- By default, `$` accepts partial matching.

- We want stricter testing, so we need to set some options.

## Our turn `"2.2.2"`

Turn off `$` partial matching when testing.

. . .

<hr>
**Only** *if needed*, `btt22::btt_reset_hard("2.2.2")`
<hr>

. . . 
 
Get new files, `btt22::btt_get("2.2.2")`:

 - `utils-testing.R`
 
`usethis::use_package("withr")` 

## `local_warn_partial_match()`

In essence, this sets (temporarily) `TRUE`:

 - `options("warnPartialMatchDollar")`
 - `options("warnPartialMatchArgs")`
 - `options("warnPartialMatchAttr")`

. . .

With a couple of wrinkles:

 - we need to treat `NULL` as `FALSE`.
 - we set things back when the *calling* scope exits.

## Our turn `"2.2.2"`, continued

- `usethis::use_test("validate")`

- add `local_warn_partial_match()` to top of blocks.

- `getOption("warnPartialMatchDollar")`

- change: 

  ```r
  expect_identical(error_condition$cols_re, "foo")
  ```

- `devtools::test()` 🎉 

- `getOption("warnPartialMatchDollar")`

- change back to `$cols_req`

## Your turn `"2.2.3"` {.smaller}

Implement testing with side effects

. . .

<hr>
**Only** *if needed*, `btt22::btt_reset_hard("2.2.3")`
<hr>

. . . 

*No* new files.

. . .

- add `local_warn_partial_match()` to all test-blocks.

- `test-matches.R`:
  - add `expect_error()` calls for data-frame and columns (use `class` argument).
  - at end, add:
   ```r
   expect_snapshot(dplyr::glimpse(italy))
   ```

## Summary

. . .

The most-common side-effect is an error. Good design:

. . .

- simple (as possible) predicate
- clear mesage
- add a `class` using naming convention, and additional information

. . .

Use snapshot tests to capture side-effects.

- be very careful when accepting changes to snapshots.

. . . 

Use `withr::local_*()` functions to "leave no footprints".

## Additional material {.smaller}

```{r}
#| eval: true
#| echo: false
vembedr::embed_youtube("ZhuHCtR3xq8", height = 400) 
```

Brian Beckman on monads (Ian watched at least ten times, learned something each time):

- tidyverse is a monoidal system

- maybe monads aren't the thing, but the functional-programming learned along the way

## References
